---
title: "ColoradoClimateAnalysis"
author: "Michael Grybko"
date: "2024-10-12"
output:
  pdf_document: default
  html_document: default
---


```{r, libraries}
library(tidyverse)
library(lubridate)
```

```{r, import climate data}
gj_data <- read.csv("USW00023066.csv")
fc_data <- read.csv("USC00053005.csv")
```

```{r, Tidy Grand Junction Data}
# Drop unnecessary columns
gj_data <- gj_data[-c(1,17:138)]

# Drop NA entries
gj_data <- gj_data %>% drop_na()

# Check for NA entries
(which(is.na(gj_data)))

# Adjust the temperature values to degrees Celsius from tenths of degrees Celsius
gj_data$TMAX <- gj_data$TMAX / 10
gj_data$TMIN <- gj_data$TMIN / 10

# Adjust the precipitaton values to millimeters from tenths of millimeters
gj_data$PRCP <- gj_data$PRCP / 10

# Convert the data column to a date format
gj_data <- gj_data %>% 
  mutate(DATE = ymd(DATE))

# Filter the data to only include years 1940 through 2023
gj_data <- gj_data %>% 
  filter(year(DATE) >= 1940 & year(DATE) <= 2023)

summary(gj_data)
```

```{r, Tidy Fort Collins Data}
# drop unnecessary columns
fc_data <- fc_data[-c(1,11,12,17:138)]

# Drop NA entries
fc_data <- fc_data %>% drop_na()

# Check for NA entries
(which(is.na(fc_data)))

# Adjust the temperature values to degrees Celsius from tenths of degrees Celsius
fc_data$TMAX <- fc_data$TMAX / 10
fc_data$TMIN <- fc_data$TMIN / 10

# Adjust the precipitaton values to millimeters from tenths of millimeters
fc_data$PRCP <- fc_data$PRCP / 10

# Convert the data column to a date format
fc_data <- fc_data %>% 
  mutate(DATE = ymd(DATE))

# Filter the data to only include years 1940 through 2023
fc_data <- fc_data %>% 
  filter(year(DATE) >= 1940 & year(DATE) <= 2023)

summary(fc_data)
```

# Precipitation by Year Grand Junction
There have been four major droughts in Colorado in the 21st century, 2002, 2012, 2018, and 2020 these will be marked with a red asterisks. Years with floods will be marked with a dark blue asterisk. 

```{r, Grand Junction Precipitation by Year}
# Sum the PRCP column by year
gj_yearly_prcp <- gj_data %>% 
  mutate(Year = year(DATE)) %>%
  group_by(Year) %>% 
  summarize(total_prcp = sum(PRCP, na.rm = TRUE), # Total precipitation by year
            ave_prcp = mean(PRCP, na.rm = TRUE))  # Average precipitation by year

# Plot total precipitation per year
gj_prcp_year <- ggplot(data = gj_yearly_prcp , aes(x = Year)) +
  geom_line(aes(y = total_prcp, color ="Total Precipitation"), size = 1.5) +
  geom_smooth(aes(y = total_prcp, color ="Smoothed Precipitation"), method ="loess", size =1, linetype=2) +  
  # Add a points for drought years with a red asterisk
  geom_point(data = gj_yearly_prcp %>% filter(Year %in% c(2002,2012,2018,2020)), 
             aes(x = Year, y = total_prcp), 
             shape = 8, size = 5, color = "red") +  
  # Add a points for regional floods with a darkblue asterisk
  geom_point(data = gj_yearly_prcp %>% filter(Year %in% c(1984,1983,1952,1957)), 
             aes(x = Year, y = total_prcp), 
             shape = 8, size = 5, color = "darkblue") +  
  scale_x_continuous(breaks = seq(1940, 2023, by = 10)) +
  scale_y_continuous(limits=c(100, 400)) +
 scale_color_manual(values = c("Total Precipitation" = "green", "Smoothed Precipitation" = "darkgreen")) +
  labs(title = "Total Precipitation per Year Grand Junction CO",
       x = "Year",
       y = "Total Precipitation in mm") +
  theme_minimal() +
  theme(plot.title = element_text(size=16, hjust = 0.5, face = "bold"),  
        axis.title = element_text(size=12, face = "bold"), 
        axis.line = element_line(size = 1, colour = "black"),
        axis.text = element_text(size=10, face="bold"), 
        legend.position = c(0.7, 0.98),  
        legend.justification = c(0, 1)) 

# save plots as jpeg
ggsave("gj_prcp_year.jpeg", width = 10, height = 6, dpi = 300)

gj_prcp_year

```

```{r, Grand Junction Snow by Year}
# Sum the PRCP column by year
gj_yearly_snow <- gj_data %>% 
  mutate(Year = year(DATE)) %>%
  group_by(Year) %>% 
  summarize(total_snow  = sum(SNOW, na.rm = TRUE), # Total precipitation by year
            ave_snow = mean(SNOW, na.rm = TRUE))  # Average precipitation by year

# Plot total snow per year
gj_snow_year <- ggplot(data = gj_yearly_snow , aes(x = Year)) +
  geom_line(aes(y = total_snow, color ="Total Snow"), size = 1.5) +
  geom_smooth(aes(y = total_snow, color ="Smoothed Snow"), method ="loess", size =1, linetype=2) +   
  # Add a points for drought years with a red asterisk
  geom_point(data = gj_yearly_snow %>% filter(Year %in% c(2002,2012,2018,2020)), 
             aes(x = Year, y = total_snow), 
             shape = 8, size = 5, color = "red") +  
  # Add a points for regional floods with a darkblue asterisk
  geom_point(data = gj_yearly_snow %>% filter(Year %in% c(1984,1983,1952,1957)), 
             aes(x = Year, y = total_snow), 
             shape = 8, size = 5, color = "darkblue") +  
  scale_x_continuous(breaks = seq(1940, 2023, by = 10)) +
 scale_color_manual(values = c("Total Snow" = "skyblue", "Smoothed Snow" = "blue")) +
  labs(title = "Total Snow per Year Grand Junction CO",
       x = "Year",
       y = "Total Snow in mm") +
  theme_minimal() +
  theme(plot.title = element_text(size=16, hjust = 0.5, face = "bold"),  
        axis.title = element_text(size=12, face = "bold"), 
        axis.line = element_line(size = 1, colour = "black"),
        axis.text = element_text(size=10, face="bold"), 
        legend.position = c(0.7, 0.98),  
        legend.justification = c(0, 1)) 

# save plots as jpegs
ggsave("gj_snow_year.jpeg", width = 10, height = 6, dpi = 300)

gj_snow_year
```

# Temperature Trends for Grand Junction
The warmest year to date in Colorado is 2012, which will be marked with an asterisk. 
To examine trends by decade the deviation from the mean maximum or minimum temperature were plotted.

```{r, Grand Junction Max Temp by Year}
gj_yearly_max_temp <- gj_data %>% 
  mutate(Year = year(DATE)) %>% 
  group_by(Year) %>% 
  summarize(ave_max_temp = mean(TMAX, na.rm = T))

# plot average max temp by year with smoothed line
gj_max_temp_year <- ggplot(data = gj_yearly_max_temp , aes(x = Year)) +
  geom_line(aes(y = ave_max_temp, color ="Average Max Temp"), size = 1.5) +
  geom_smooth(aes(y = ave_max_temp, color ="Smoothed Temp"), method ="loess", size = 1, linetype=2) +  
  # Add a point for 2012 with an asterisk
  geom_point(data = gj_yearly_max_temp %>% filter(Year == 2012), 
             aes(x = Year, y = ave_max_temp), 
             shape = 8, size = 5, color = "black") +  # Asterisk for 2012
  scale_x_continuous(breaks = seq(1940, 2023, by = 10)) +
  scale_y_continuous(limits=c(16, 21)) +
  scale_color_manual(values = c("red", "darkred"), 
                    labels = c("Average Max Temp", "Smoothed Max Temp")) +
  labs(title = "Average Max Temperature by Year Grand Junction CO",
       x = "Year",
       y = "Max Temp °C") +
  theme_minimal() +
  theme(plot.title = element_text(size=16, hjust = 0.5, face = "bold"),  
        axis.title = element_text(size=12, face = "bold"), 
        axis.line = element_line(size = 1, colour = "black"),
        axis.text = element_text(size=10, face="bold"), 
        legend.position = c(0.025, 0.95),  
        legend.justification = c(0, 1)) 

ggsave("gj_max_temp_year.jpeg", width = 10, height = 6, dpi = 300)
gj_max_temp_year
```

```{r, Grand Junction Max Temp by Decade}
# Calculate the mean of TMAX
gj_ave_max_temp <- mean(gj_data$TMAX)
# Group by decade and calculate the difference from mean
gj_decade_max_temp <- gj_data %>% 
  mutate(Year = year(DATE),
         Decade = (Year %/% 10) * 10) %>% 
  filter(year(DATE) < 2020) %>% 
  group_by(Decade) %>% 
  summarize(
    diff_ave_max_temp = (mean(TMAX, na.rm = TRUE) - gj_ave_max_temp),
    sd_max_temp = sd(TMAX, na.rm = TRUE),   # Standard deviation of max temperature
    count = n()                            # Number of observations per decade
  )
# Compute standard error 
gj_decade_max_temp<- gj_decade_max_temp %>%
  mutate(se_max_temp = sd_max_temp / sqrt(count))

# Categorize each decade as positive or negative
gj_decade_max_temp <- gj_decade_max_temp %>%
  mutate(temp_category = ifelse(diff_ave_max_temp >= 0, "Positive", "Negative"))


# Create the bar plot with color gradient based on the value of diff_ave_max_temp
gj_max_temp_decade <- ggplot(data = gj_decade_max_temp, 
                             aes(x = as.factor(Decade), 
                                 y = diff_ave_max_temp, 
                                 fill = diff_ave_max_temp)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = diff_ave_max_temp - se_max_temp, 
                    ymax = diff_ave_max_temp + se_max_temp), 
                width = 0.2, color = "black") +
  scale_fill_gradient2(low = "skyblue1", mid = "white", high = "firebrick2", 
                       midpoint = 0, 
                       name = "Temp Deviation") +
  labs(title = "Changes in Average Max Temperature by Decade in Grand Junction",
       x = "Decade",
       y = "Deviation from Average Max Temperature (°C)") +
  theme_minimal() +
  theme(plot.title = element_text(size=14, hjust = 0.5, face = "bold"), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    axis.line = element_line(size = 1, colour = "black")
  )

# Save the updated plot as a JPEG file
ggsave("gj_max_temp_decade_anomalies_scaled.jpeg", plot = gj_max_temp_decade, width = 10, height = 6, dpi = 300)

# Display the plot
gj_max_temp_decade
```

```{r, Grand Junction Min Temp by Year}
gj_yearly_min_temp <- gj_data %>% 
  mutate(Year = year(DATE)) %>% 
  group_by(Year) %>% 
  summarize(ave_min_temp = mean(TMIN, na.rm = TRUE))

gj_min_temp_year <- ggplot(data = gj_yearly_min_temp , aes(x = Year)) +
  geom_line(aes(y = ave_min_temp, color = "Average Min Temp"), size = 1.5) +
  geom_smooth(aes(y = ave_min_temp, color ="Smoothed Temp"), method ="loess", size = 1, linetype=2) + 
  # Add a point for 2012 with an asterisk
  geom_point(data = gj_yearly_min_temp %>% filter(Year == 2012), 
             aes(x = Year, y = ave_min_temp), 
             shape = 8, size = 5, color = "black") +  # Asterisk for 2012
  scale_x_continuous(breaks = seq(1940, 2023, by = 10)) +
  scale_color_manual(values = c("blue", "navyblue"), 
                    labels = c("Average Min Temp", "Smoothed Min Temp")) +
  labs(title = "Average Minimum Temperature by Year Grand Junction CO",
       x = "Year",
       y = "Average Min Temp °C") +
  theme_minimal() +
  theme(plot.title = element_text(size=16, hjust = 0.5, face = "bold"),  
        axis.title = element_text(size=12, face = "bold"), 
        axis.line = element_line(size = 1, colour = "black"),
        axis.text = element_text(size=10, face="bold"),  
        legend.position = c(0.75, 0.95),  
        legend.justification = c(0, 1)) 

ggsave("gj_min_temp_year.jpeg", width = 10, height = 6, dpi = 300)
gj_min_temp_year
```

```{r, Grand Junction Min Temp by Decade}
# Calculate the mean of TMIN
gj_ave_min_temp <- mean(gj_data$TMIN)
# Group by decade and calculate the difference from mean
gj_decade_min_temp <- gj_data %>% 
  mutate(Year = year(DATE),
         Decade = (Year %/% 10) * 10) %>% 
  filter(year(DATE) < 2020) %>% 
  group_by(Decade) %>% 
  summarize(
    diff_ave_min_temp = (mean(TMIN, na.rm = TRUE) - gj_ave_min_temp),
    sd_min_temp = sd(TMIN, na.rm = TRUE),   # Standard deviation of max temperature
    count = n()                            # Number of observations per decade
  )
# Compute standard error 
gj_decade_min_temp<- gj_decade_min_temp %>%
  mutate(se_min_temp = sd_min_temp / sqrt(count))

# Categorize each decade as positive or negative
gj_decade_min_temp <- gj_decade_min_temp %>%
  mutate(temp_category = ifelse(diff_ave_min_temp >= 0, "Positive", "Negative"))

# Create the bar plot with color gradient based on the value of diff_ave_max_temp
gj_min_temp_decade <- ggplot(data = gj_decade_min_temp, 
                             aes(x = as.factor(Decade), 
                                 y = diff_ave_min_temp, 
                                 fill = diff_ave_min_temp)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = diff_ave_min_temp - se_min_temp, 
                    ymax = diff_ave_min_temp + se_min_temp), 
                width = 0.2, color = "black") +
  scale_fill_gradient2(low = "skyblue1", mid = "white", high = "firebrick2", 
                       midpoint = 0, 
                       name = "Temp Deviation") +
  labs(title = "Changes in Average Min Temperature by Decade in Grand Junction",
       x = "Decade",
       y = "Deviation from Average Min Temperature (°C)") +
  theme_minimal() +
  theme(plot.title = element_text(size=14, hjust = 0.5, face = "bold"), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    axis.line = element_line(size = 1, colour = "black")
  )

# Save the bar plot as a JPEG file
ggsave("gj_min_temp_decade.jpeg",width = 10, height = 6, dpi = 300)

(gj_min_temp_decade)
```
# Precipitation by Year For Fort Collins
There have been four major droughts in Colorado in the 21st century, 2002, 2012, 2018, and 2020 these will be marked with a red asterisks. Years with floods will be marked with a dark blue asterisk. 

```{r, Fort Collins Precipitation by Year}
# Sum the PRCP column by year
fc_yearly_prcp <- fc_data %>% 
  mutate(Year = year(DATE)) %>%
  group_by(Year) %>% 
  summarize(total_prcp = sum(PRCP, na.rm = TRUE), # Total precipitation by year
            ave_prcp = mean(PRCP, na.rm = TRUE))  # Average precipitation by year

# Plot total precipitation per year
fc_prcp_year <- ggplot(data = fc_yearly_prcp , aes(x = Year)) +
  geom_line(aes(y = total_prcp, color ="Total Precipitation"), size = 1.5) +
  geom_smooth(aes(y = total_prcp, color ="Smoothed Precipitation"), method ="loess", size =1, linetype=2) +  
  # Add a points with an asterisk
  geom_point(data = fc_yearly_prcp %>% filter(Year %in% c(2002,2012,2018,2020)), 
             aes(x = Year, y = total_prcp), 
             shape = 8, size = 5, color = "red") + 
  scale_x_continuous(breaks = seq(1940, 2023, by = 10)) +
  # Add a points for regional floods with a darkblue asterisk
  geom_point(data = fc_yearly_prcp %>% filter(Year %in% c(2013,1997)), 
             aes(x = Year, y = total_prcp), 
             shape = 8, size = 5, color = "darkblue") + 
  #scale_y_continuous(limits=c(100, 400)) +
 scale_color_manual(values = c("Total Precipitation" = "green", "Smoothed Precipitation" = "darkgreen")) +
  labs(title = "Total Precipitation per Year Fort Collins CO",
       x = "Year",
       y = "Total Precipitation in mm") +
  theme_minimal() +
  theme(plot.title = element_text(size=16, hjust = 0.5, face = "bold"),  
        axis.title = element_text(size=12, face = "bold"), 
        axis.line = element_line(size = 1, colour = "black"),
        axis.text = element_text(size=10, face="bold"), 
        legend.position = c(0.7, 0.98),  
        legend.justification = c(0, 1)) 

# save plot as jpeg
ggsave("fc_prcp_year.jpeg", width = 10, height = 6, dpi = 300)

fc_prcp_year

```

```{r, Fort Colllins Snow by Year}
# Sum the PRCP column by year
fc_yearly_snow <- fc_data %>% 
  mutate(Year = year(DATE)) %>%
  group_by(Year) %>% 
  summarize(total_snow = sum(SNOW, na.rm = TRUE), # Total precipitation by year
            ave_snow = mean(SNOW, na.rm = TRUE))  # Average precipitation by year

# Plot total precipitation per year
fc_snow_year <- ggplot(data = fc_yearly_snow , aes(x = Year)) +
  geom_line(aes(y = total_snow, color ="Total Snow"), size = 1.5) +
  geom_smooth(aes(y = total_snow, color ="Smoothed Snow"), method ="loess", size =1, linetype=2) + 
  # Add a points with an asterisk
  geom_point(data = fc_yearly_snow %>% filter(Year %in% c(2002,2012,2018,2020)), 
             aes(x = Year, y = total_snow), 
             shape = 8, size = 5, color = "red") + 
  # Add a points for regional floods with a darkblue asterisk
  geom_point(data = fc_yearly_snow %>% filter(Year %in% c(2013,1997)), 
             aes(x = Year, y = total_snow), 
             shape = 8, size = 5, color = "darkblue") + 
  scale_x_continuous(breaks = seq(1940, 2023, by = 10)) +
 scale_color_manual(values = c("Total Snow" = "skyblue", "Smoothed Snow" = "blue")) +
  labs(title = "Total Snow per Year Fort Collins CO",
       x = "Year",
       y = "Total Snow in mm") +
  theme_minimal() +
  theme(plot.title = element_text(size=16, hjust = 0.5, face = "bold"),  
        axis.title = element_text(size=12, face = "bold"), 
        axis.line = element_line(size = 1, colour = "black"),
        axis.text = element_text(size=10, face="bold"), 
        legend.position = c(0.7, 0.98),  
        legend.justification = c(0, 1)) 

# save plots as jpegs
ggsave("fc_snow_year.jpeg", width = 10, height = 6, dpi = 300)

fc_snow_year
```

# Temperature Trends for Fort Collins
The warmest year to date in Colorado is 2012, which will be marked with an asterisk. 
To examine trends by decade the deviation from the mean maximum or minimum temperature were plotted.

```{r, Fort Collins Max Temp by Year}
fc_yearly_max_temp <- fc_data %>% 
  mutate(Year = year(DATE)) %>% 
  group_by(Year) %>% 
  summarize(ave_max_temp = mean(TMAX, na.rm = T))

# plot average max temp by year with smoothed line
fc_max_temp_year <- ggplot(data = fc_yearly_max_temp , aes(x = Year)) +
  geom_line(aes(y = ave_max_temp, color ="Average Max Temp"), size = 1.5) +
  geom_smooth(aes(y = ave_max_temp, color ="Smoothed Temp"), method ="loess", size = 1, linetype=2) +   
  # Add a point for 2012 with an asterisk
  geom_point(data = fc_yearly_max_temp %>% filter(Year == 2012), 
             aes(x = Year, y = ave_max_temp), 
             shape = 8, size = 5, color = "black") +  # Asterisk for 2012
  scale_x_continuous(breaks = seq(1940, 2023, by = 10)) +
  scale_y_continuous(limits=c(15, 21)) +
  scale_color_manual(values = c("red", "darkred"), 
                    labels = c("Average Max Temp", "Smoothed Max Temp")) +
  labs(title = "Average Max Temperature by Year Fort Collins CO",
       x = "Year",
       y = "Max Temp °C") +
  theme_minimal() +
  theme(plot.title = element_text(size=16, hjust = 0.5, face = "bold"),  
        axis.title = element_text(size=12, face = "bold"), 
        axis.line = element_line(size = 1, colour = "black"),
        axis.text = element_text(size=10, face="bold"), 
        legend.position = c(0.005, 0.995),  
        legend.justification = c(0, 1)) 

ggsave("fc_max_temp_year.jpeg", width = 10, height = 6, dpi = 300)
fc_max_temp_year
```

```{r, Fort Collins Max Temp by Decade}
# Calculate the mean of TMAX
fc_ave_max_temp <- mean(fc_data$TMAX)
# Group by decade and calculate the difference from mean
fc_decade_max_temp <- fc_data %>% 
  mutate(Year = year(DATE),
         Decade = (Year %/% 10) * 10) %>% 
  filter(year(DATE) < 2020) %>% 
  group_by(Decade) %>% 
  summarize(
    diff_ave_max_temp = (mean(TMAX, na.rm = TRUE) - fc_ave_max_temp),
    sd_max_temp = sd(TMAX, na.rm = TRUE),   # Standard deviation of max temperature
    count = n()                            # Number of observations per decade
  )
# Compute standard error 
fc_decade_max_temp<- fc_decade_max_temp %>%
  mutate(se_max_temp = sd_max_temp / sqrt(count))

# Categorize each decade as positive or negative
fc_decade_max_temp <- fc_decade_max_temp %>%
  mutate(temp_category = ifelse(diff_ave_max_temp >= 0, "Positive", "Negative"))

# Create the bar plot with color gradient based on the value of diff_ave_max_temp
fc_max_temp_decade <- ggplot(data = fc_decade_max_temp, 
                             aes(x = as.factor(Decade), 
                                 y = diff_ave_max_temp, 
                                 fill = diff_ave_max_temp)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = diff_ave_max_temp - se_max_temp, 
                    ymax = diff_ave_max_temp + se_max_temp), 
                width = 0.2, color = "black") +
  scale_fill_gradient2(low = "skyblue1", mid = "white", high = "firebrick2", 
                       midpoint = 0, 
                       name = "Temp Deviation") +
  labs(title = "Changes in Average Max Temperature by Decade in Fort Collins",
       x = "Decade",
       y = "Deviation from Average Max Temperature (°C)") +
  theme_minimal() +
  theme(plot.title = element_text(size=14, hjust = 0.5, face = "bold"), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    axis.line = element_line(size = 1, colour = "black")
  )
# Save the bar plot as a JPEG file
ggsave("fc_max_temp_decade.jpeg",width = 10, height = 6, dpi = 300)

(fc_max_temp_decade)
```

```{r, Fort Collins Min Temp by Year}
fc_yearly_min_temp <- fc_data %>% 
  mutate(Year = year(DATE)) %>% 
  group_by(Year) %>% 
  summarize(ave_min_temp = mean(TMIN, na.rm = TRUE))

fc_min_temp_year <- ggplot(data = fc_yearly_min_temp , aes(x = Year)) +
  geom_line(aes(y = ave_min_temp, color = "Average Min Temp"), size = 1.5) +
  geom_smooth(aes(y = ave_min_temp, color ="Smoothed Temp"), method ="loess", size = 1, linetype=2) + 
  # Add a point for 2012 with an asterisk
  geom_point(data = fc_yearly_min_temp %>% filter(Year == 2012), 
             aes(x = Year, y = ave_min_temp), 
             shape = 8, size = 5, color = "black") +  # Asterisk for 2012
  scale_x_continuous(breaks = seq(1940, 2023, by = 10)) +
  scale_color_manual(values = c("blue", "navyblue"), 
                    labels = c("Average Min Temp", "Smoothed Min Temp")) +
  labs(title = "Average Minimum Temperature by Year Fort Collins CO",
       x = "Year",
       y = "Min Temp °C") +
  theme_minimal() +
  theme(plot.title = element_text(size=16,hjust = 0.5, face = "bold"),  
        axis.title = element_text(size=12, face = "bold"), 
        axis.line = element_line(size = 1, colour = "black"),
        axis.text = element_text(size=10, face="bold"), 
        legend.position = c(0.025, 0.95),  
        legend.justification = c(0, 1)) 

ggsave("fc_min_temp_year.jpeg", width = 10, height = 6, dpi = 300)
fc_min_temp_year
```

```{r, Fort Collins Min Temp by Decade}
# Calculate the mean of TMIN
fc_ave_min_temp <- mean(fc_data$TMIN)
# Group by decade and calculate the difference from mean
fc_decade_min_temp <- fc_data %>% 
  mutate(Year = year(DATE),
         Decade = (Year %/% 10) * 10) %>% 
  filter(year(DATE) < 2020) %>% 
  group_by(Decade) %>% 
  summarize(
    diff_ave_min_temp = (mean(TMIN, na.rm = TRUE) - fc_ave_min_temp),
    sd_min_temp = sd(TMIN, na.rm = TRUE),   # Standard deviation of max temperature
    count = n()                            # Number of observations per decade
  )
# Compute standard error 
fc_decade_min_temp<- fc_decade_min_temp %>%
  mutate(se_min_temp = sd_min_temp / sqrt(count))

# Categorize each decade as positive or negative
fc_decade_min_temp <- fc_decade_min_temp %>%
  mutate(temp_category = ifelse(diff_ave_min_temp >= 0, "Positive", "Negative"))

# Create the bar plot with color gradient based on the value of diff_ave_max_temp
fc_min_temp_decade <- ggplot(data = fc_decade_min_temp, 
                             aes(x = as.factor(Decade), 
                                 y = diff_ave_min_temp, 
                                 fill = diff_ave_min_temp)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = diff_ave_min_temp - se_min_temp, 
                    ymax = diff_ave_min_temp + se_min_temp), 
                width = 0.2, color = "black") +
  scale_fill_gradient2(low = "skyblue1", mid = "white", high = "firebrick2", 
                       midpoint = 0, 
                       name = "Temp Deviation") +
  labs(title = "Changes in Average Min Temperature by Decade in Fort Collins",
       x = "Decade",
       y = "Deviation from Average Min Temperature (°C)") +
  theme_minimal() +
  theme(plot.title = element_text(size=14, hjust = 0.5, face = "bold"), 
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.title = element_text(size = 12, face = "bold"),
    axis.text = element_text(size = 10, face = "bold"),
    axis.line = element_line(size = 1, colour = "black")
  )

# Save the bar plot as a JPEG file
ggsave("fc_min_temp_decade.jpeg",width = 10, height = 6, dpi = 300)

(fc_min_temp_decade)
```

```{r, Fort Collins Max Temp by Season}
fc_seasonal_max_temp <- fc_data %>% 
  mutate(Year = year(DATE), 
         Month = month(DATE), 
         Season = case_when(
           Month %in% c(12, 1, 2) ~ "Winter",
           Month %in% c(3, 4, 5) ~ "Spring",
           Month %in% c(6, 7, 8) ~ "Summer",
           Month %in% c(9, 10, 11) ~ "Fall"
         )) %>% 
  group_by(Year, Season) %>% 
  summarize(ave_max_temp = mean(TMAX, na.rm = TRUE))

fc_fall_max_temp <- fc_seasonal_max_temp %>% 
  filter(Season == "Fall")

# plot average max temp by month with smoothed line
fc_temp_plot_fall <- ggplot(data = fc_fall_max_temp, aes(x = Year)) +
  geom_line(aes(y = ave_max_temp, color = "Average Max Temp"), size = 1.5) +
  geom_smooth(aes(y = ave_max_temp, color ="Smoothed Temp"), method ="loess", size = 1, linetype=2) +   
  scale_x_continuous(breaks = seq(1940, 2023, by = 10)) +
  scale_color_manual(values = c("red", "darkred"), 
                    labels = c("Average Max Temp", "Smoothed Max Temp")) + 
  labs(title = "Average Fall Max Temperature Fort Collins CO",
       x = "Year",
       y = "Average Max Temperature (°C)") +  
  theme_minimal() +
  theme(plot.title = element_text(size=16,hjust = 0.5, face = "bold"),  
        axis.title = element_text(size=12, face = "bold"), 
        axis.line = element_line(size = 1, colour = "black"),
        axis.text = element_text(size=10, face="bold"), 
        legend.position = c(0.005, 0.995),  
        legend.justification = c(0, 1),
        panel.grid.minor = element_blank())

ggsave("fc_temp_plot_fall.jpeg", plot = fc_temp_plot_fall, width = 10, height = 6, dpi = 300)

fc_temp_plot_fall
```

# Linear Models
Simple linear models with either maximum or minimum temperature as the response and date as the predictor for Grand Junction and Fort Collins. All models show a significant relationship between temperature and date accept the the minimum temperature data for Grand Junction.

```{r, linear model Grand Junction maximum temp}
gj.max.lm <- lm(ave_max_temp ~ Year, data = gj_yearly_max_temp)
summary(gj.max.lm)
```

```{r, linear model Grand Junction minimum temp}
gj.min.lm <- lm(ave_min_temp ~ Year, data = gj_yearly_min_temp)
summary(gj.min.lm)
```

```{r, linear model Fort Collins maximum temp}
fc.max.lm <- lm(ave_max_temp ~ Year, data = fc_yearly_max_temp)
summary(fc.max.lm)
```

```{r, linear model Fort Collins minimum temp}
fc.min.lm <- lm(ave_min_temp ~ Year, data = fc_yearly_min_temp)
summary(fc.min.lm)
```

